Predicting the (non-cryptographic) PCG Pseudo-Random Number Generator In Practice

### 2020-06-01 article/main.tex
section 5 en cours

\subsection{Reconstruction in ``High'' Dimension Using an Exact CVP Solver}

@@

\subsection{Reconstruction in Low Dimension Using Babai's Rounding}

@@

We know $K_i[58:64+\ell]$ for all $i$, and for each guessed rotation $r_i$ we
have access to $T_i \stackrel{def}{=} S_i[58:64+\ell]$.  We want
$T'_i \stackrel{def}{=} (S_i \boxminus K_i)[58:64+\ell]$, which is the
truncation of $S'_i$. Thanks to~\eqref{eq:borrow}, we know that there is an
unknown vector $B$ of \emph{borrows}, whose components are either 0 or 1, such
that $T' = S[58:64+\ell] \boxminus K[58:64+\ell] \boxminus B$. Because the
borrows are unknown, we in fact compute
$\widetilde{T'} = S[58:64+\ell] \boxminus K[58:64+\ell]$, and clearly
$\widetilde{T'} = T' \boxplus B$. We are thus in the context of the problem
discussed in section~\ref{sec:geometric}, namely reconstructing a geometric
sequence given $6+\ell$ (noisy) most-significant bits. The ``noise'' is the
unknown vector $B$ of borrows.

@@

The algorithm that reconstructs the internal state of the \textsf{PCG64}
generator with known increment proceeds as shown in
algorithm~\ref{algo:known}. The point is that when the guesses are correct, then
from the truncated geometric series $\mathbf{T'}$, the solution of the CVP
instance reveals $U_j = S_j'[\ell:64+\ell]$. From there, the correction of the
algorithm is easily established.

@@

@@

A few details still need to be fleshed out. To be precise, let us assume that we
have guessed the $\ell$ least-significant bits of $S_0$ (we denote them by
$w_0$) and the first rotation $r_0$. Set $Y_0 = X_0 \lll r_0$. % there is an
% unknown value $x_0$ such that:
% \[
%   S_0 = w_0 + 2^\ell x + 2^{64} (w_0 \oplus Y_0[0:\ell]) + 2^{64+\ell} (x \oplus Y_0[\ell:64]) \qquad \left(0 \leq x < 2^{64-\ell}\right).
% \]
%The unknown value $x_0$ will not be recovered until the third phase of the whole
%procedure.
We obtain the $i$-th state by $S_i \equiv \nabla S_i \boxplus S_0$; however,
because the ``middle'' of $S_0$ is unknown, then an unknown \emph{carry} may
cross the 64-th bit during the addition and perturb $S_i[64:64+\ell]$. As a
result, there is an unknown vector $C$, whose components are either 0 or 1, such
that such that:
\begin{align*}
  S_i[64:64+\ell] &= C_i \boxplus \nabla S_i[64:64+\ell] \boxplus (w_0 \oplus Y_0[0:\ell]) \\
\end{align*}

@@

\begin{algorithm}
\begin{algorithmic}[1]
  \Procedure{ConsistencyCheck}{$\Delta S_0, w_0, Y_0, X_5, \dots, X_k$}
  \State $v_0 = w_0 \oplus Y_0[0:\ell]$ \Comment{$v_0 = S_0[64:64+\ell]$}
  \For{$i=5, \dots, k$}
  \State $u_i \gets \Delta S_0 (a^i-1)(a-1)^{-1} \bmod 2^{64+\ell}$ \Comment{$u_i = \nabla S_i[0:64+\ell]$}
  \State $w_i = w_0 + u_i[0:\ell] \bmod 2^{\ell}$ \Comment{$w_i = S_i[0:\ell]$}
  \State $v_i = v_0 + u_i[64:64+\ell] \bmod 2^{\ell}$ \Comment{$S_i[64:64+\ell] \in \{v_i, v'_i\}$}
  \State $v'_i = v_i + 1 \bmod 2^{\ell}$
  \State $\mathcal{C}_i \gets \{ w_i \oplus (X_i \lll r_i)[0:\ell]~|~ 0\leq r_i < 64\}$ \Comment{Check eq.~\eqref{eq:consistency}}
  \If{$\{v_i, v'_i\} \cap \mathcal{C}_i = \emptyset$}
  \State \Return \textbf{False} \Comment{Bad Guesses}
  \EndIf
  \EndFor
  \State \Return \textbf{True} \Comment{No inconsistency}
  \EndProcedure

@@

\Procedure{ReconstructPartialDifference}{$X_0, \dots, X_k$}
  \For{$0 \leq w_0 < 2^{\ell}$ and $0 \leq c < 2^{\ell - 1}$} \Comment{Guess least-significant bits}
  \State $K_j \gets a^j w_0 + (2c+1)(a^j - 1)(a-1)^{-1} \bmod 2^{128}$ \Comment{Known part}
  \For{$0 \leq r_0, r_1, r_2, r_3, r_4 < 64$} \Comment{Guess rotations}
  \State $Y_j \gets X_j \lll r_j$ \Comment{Undo rotations}
  \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(K_j \oplus Y_j\right)[0:\ell]$ \Comment{Truncated LCG}
  \State $T'_j \gets T_j \boxminus  K_i[58:64+\ell]$ \Comment{Truncated geometric series}
  \State \todo{Attention double retenue : $\pm 1$.}
  \State $\Delta T'_j = (T'_{j+1} - T'_j) \bmod 2^{6 + \ell}$  \Comment{Difference}
  \State $(\Delta U_0, \dots, \Delta U_3) \gets \left\lfloor (\Delta T'_0, \dots, \Delta T'_3) \cdot 2^{58-\ell} \cdot \widetilde G_4^{-1} \right\rceil \cdot \widetilde G_4$ \Comment{CVP}
  \State $\Delta S_0[0:64+\ell] \gets \left((w_1 - w_0) \bmod 2^{\ell}\right) + 2^{\ell} \cdot \Delta U_0[0:58-\ell]$ \Comment{Check}
  \If{$\Call{ConsistencyCheck}{\Delta_0, w_0, Y_0, X_5, \dots, X_k}$}
  \State \Return $(w_0, c, r_0, \dots, r_4, \Delta_0)$.
  \EndIf
  \EndFor
  \EndFor
  \EndProcedure
\end{algorithmic}
\caption{Partial difference reconstruction algorithm (when $c$ is unknown).}
\label{algo:unknown_1}
\end{algorithm}

The heart of the algorithm is again the reconstruction of a truncated geometric
progression. Looking at table~\ref{tab:babai}, we see that the best choice
consists in guessing 5 consecutive rotations and $\ell=13$ least-significant
bits. Therefore, \textsc{ReconstructPartialDifference} does $2^{55}$ iterations
of the inner loop, and succeeds deterministically.

% \begin{theorem}
%   If $\ell \geq 13$, then algorithm \textsc{ReconstructPartialDifference} is
%   correct (it always returns the correct values).
% \end{theorem}

% \begin{proof}
%   \todo{WRITE-ME}
%   \begin{align*}
%   \mathbf{S} &= \LCG(S_0, c) \\
%              &= \LCG\left(S_0[0:\ell] + 2^\ell S_0[\ell:128], c[0:\ell] + 2^\ell c[\ell:128]\right) \\
%              &= \LCG\left(S_0[0:\ell], c[0:\ell]\right) + 2^\ell \LCG\left(S_0[\ell:128],  c[\ell:128]\right) \\
%   \Delta \mathbf{S} &= \LCG(S_1, c) - \LCG(S_0, c) \\
%              &= \LCG(S_1 - S_0, 0)
% \end{align*}
% \end{proof}

Knowing the values of $\Delta S_0[0:64+\ell]$ as well as the $\ell$
least-significant bits of $S_0$ and $c$ is sufficient to get rid the nastier
feature \textsf{PCG64}: armed with this knowledge, we can determine all the
subsequent rotations deterministically, at negligible cost, using
eq~\eqref{eq:find_rotation}.

For each $i$, it suffices to try the $64$ possible values of $r_i$; only one
should satisfy the equation. Using $X_0, \dots, X_{40}$, we thus recover the 6
most-significant bits of $S_0, \dots, S_{40}$. We therefore compute the 6 most
significant bits of the differences $\Delta S_i$ between consecutive states (up
to missing carries), and we are faced with the problem of reconstructing a
128-bit geometric progression using 39 consecutive outputs truncated to their 6
most-significant bits. This can be done by solving an instance of CVP in
dimension $\leq 40$. We use the off-the-shelf CVP solver embedded in
\texttt{fplll}: it runs in no measurable time.

More precisely, there are still details to be dealt with regarding
carries. Indeed, the \textsc{ReconstructRotations} returns the 6 most
significant bits of any number of $S_i$'s (denoted by $r_i$). Therefore, there
are unknown 122-bit quantities $x_i$ such that:
\begin{align*}
  S_i &= 2^{122} r_i + x_i \\
  \Delta S_i &\equiv 2^{122} (r_{i+1} - r_i) + (x_{i+1} - x_{i}) \bmod 2^{128}
\end{align*}

\todo{check indices nabla}

@@

  %\State $x_i = x_0 + \left(\Delta S_0 \frac{a^i-1}{a-1} \bmod 2^{64+\ell}\right)[64:64+\ell]$. \Comment{$x_i = S_i[64:64+\ell]$}
  \For{$0 \leq r < 64$} \Comment{Try all rotations}
  \If{$w_i \oplus (X_i \lll r)[0:\ell] \in \left\{v_i, v'_i\right\}$} \Comment{Check eq.~\eqref{eq:find_rotation}}
  \State $\mathcal{H} \gets r \texttt{::} \mathcal{H}$ \Comment{New candidate $r_i$}
  \EndIf
  \EndFor
  \State \Return $\left\{ h\texttt{::} t~|~ h \in \mathcal{H}, t \in \mathcal{T} \right\}$ \Comment{Return $\mathcal{H} \times \mathcal{T}$}
  \EndFunction
  \State
  \Function{ReconstructFullDifference}{...}
  \State \todo{WRITEME}
  \EndFunction
\end{algorithmic}
\caption{Rotations and full difference reconstruction algorithm}
\label{algo:unknown_2}
\end{algorithm}

% \begin{theorem}
%   The algorithm described above works with probability.... in time ...
% \end{theorem}

>>>>

\subsection{Reconstruction in ``High'' Dimension Using an Exact CVP Solver}
\label{sec:cvp_big}

@@

\subsection{Reconstruction in Low Dimension Using Babai's Rounding}
\label{sec:cvp_small}

@@

We know $K_i[58:64+\ell]$ for all $i$, and for each guessed rotation $r_i$ we
have access to $T_i \stackrel{def}{=} S_i[58:64+\ell]$.  We want
$T'_i \stackrel{def}{=} (S_i \boxminus K_i)[58:64+\ell]$, which is the
truncation of $S'_i$. Thanks to~\eqref{eq:borrow}, we know that there is an
unknown vector $B$ of \emph{borrows}, whose components are either 0 or 1, such
that $T' = S[58:64+\ell] \boxminus K[58:64+\ell] \boxminus B$. Because the
borrows are unknown, we in fact compute
$\widetilde{T'} = S[58:64+\ell] \boxminus K[58:64+\ell]$, and clearly
$\widetilde{T'} = T' \boxplus B$. We are thus in the context of the problem
discussed in section~\ref{sec:geometric}, namely reconstructing a geometric
sequence given $6+\ell$ (noisy) most-significant bits. The ``noise'' is the
unknown vector $B$ of borrows. %Notice that the noise is asymmetric.

@@

The algorithm that reconstructs the internal state of the \textsf{PCG64}
generator with known increment proceeds as shown in
algorithm~\ref{algo:known}. The point is that when the guesses are correct, then
from the truncated geometric series $T'$, the solution of the CVP instance
reveals $U_j = S_j'[\ell:64+\ell]$. From there, the correction of the algorithm
is easily established.

@@

We consider again the sequence of internal states
$S = (S_0, S_1, \dots) = \LCG_{128}(S_0, c)$. We will guess the $\ell$
least-significant bits of $S_0$ and of $c$, therefore let us assume that their
value is known and denote it by $w_0$ and $c_0$. We define
$S' = \LCG_{128}(S_0 - w, c - c_0)$ and $K = \LCG_{128}(w_0, c_0)$ --- again,
$K$ is Known and $S' = S - K$. This time, the components of $S'$ do \emph{not}
follow a geometric progression; but we still have that the $\ell$ least
significant bits of each $S'_i$ are zero. Set
$\Delta S'_i \stackrel{def}{=} S'_{i+1} - S'_i$; $\Delta S'[\ell:64+\ell]$
follows a geometric progression of common ratio $a$ modulo $2^{64}$
(again). This time, we have to find $\Delta S'_0[\ell:64+\ell]$.

As in section~\ref{sec:Cknown}, we have access to
$T_i \stackrel{def}{=} S_i[58:64+\ell]$. We want to subtract the known part to
obtain $T'_i \stackrel{def}{=} (S_i \boxminus K_i)[58:64+\ell]$, which is the
truncation of $S'_i$. This again introduces an unknown vector $B$ of
\emph{borrows}, and in fact we can only compute
$\widetilde{T'} = S[58:64+\ell] \boxminus K[58:64+\ell]$, with
$\widetilde{T'} = T' \boxplus B$. As explained above, to access a geometric
sequence, we would like to obtain
$\Delta T'_i \stackrel{def}{=} T'_{i+1} - T'_i$, but we can only compute:
\[
  \Delta \widetilde{T}'_i \stackrel{def}{=} \widetilde{T}'_{i+1} - \widetilde{T}'_i
  =  (T'_{i+1} \boxminus T'_{i}) \boxplus (B_{i+1} \boxminus B_{i})
\]

We are thus still in the context of the problem discussed in
section~\ref{sec:geometric}, but this time the ``noise'' caused by the carries
is given by $B_{i+1} - B_i$. When the guesses are correct, then Babai's rounding
will reconstruct $\Delta \widetilde{S}'[\ell:64+\ell]$ from
$\Delta \widetilde{T}'$. This in turn yields $\Delta S_0[0:64+\ell]$.

@@

A few details still need to be fleshed out. To be precise, let us assume that we
have guessed the $\ell$ least-significant bits of $S_0$ (we denote them by
$w_0$) and the first rotation $r_0$. Set $Y_0 = X_0 \lll r_0$. % there is an
% unknown value $x_0$ such that:
% \[
%   S_0 = w_0 + 2^\ell x + 2^{64} (w_0 \oplus Y_0[0:\ell]) + 2^{64+\ell} (x \oplus Y_0[\ell:64]) \qquad \left(0 \leq x < 2^{64-\ell}\right).
% \]
%The unknown value $x_0$ will not be recovered until the third phase of the whole
%procedure.
We obtain the $i$-th state by $S_i \equiv \nabla S_i \boxplus S_0$; however,
because the ``middle'' of $S_0$ is unknown, then an unknown \emph{carry} may
cross the 64-th bit during the addition and perturb $S_i[64:64+\ell]$. As a
result, there is an unknown vector $C$, whose components are either 0 or 1, such
that such that:
\begin{align*}
  S_i[64:64+\ell] &= C_i \boxplus \nabla S_i[64:64+\ell] \boxplus (\underbrace{w_0 \oplus Y_0[0:\ell]}_{S_0[64:64+\ell]}) \\
\end{align*}

@@

\begin{algorithm}
\begin{algorithmic}[1]
  \Procedure{ConsistencyCheck}{$\Delta S_0, w_0, Y_0, X_5, \dots, X_k$}
  \State $v_0 = w_0 \oplus Y_0[0:\ell]$ \Comment{$v_0 = S_0[64:64+\ell]$}
  \For{$i=5, \dots, k$}
  \State $u_i \gets \Delta S_0 (a^i-1)(a-1)^{-1} \bmod 2^{64+\ell}$ \Comment{$u_i = \nabla S_i[0:64+\ell]$}
  \State $w_i = w_0 \boxplus u_i[0:\ell]$ \Comment{$w_i = S_i[0:\ell]$}
  \State $v_i = v_0 \boxplus u_i[64:64+\ell]$ \Comment{$S_i[64:64+\ell] \in \{v_i, v'_i\}$}
  \State $v'_i = v_i \boxplus 1$
  \State $\mathcal{C}_i \gets \{ w_i \oplus (X_i \lll r_i)[0:\ell]~|~ 0\leq r_i < 64\}$ \Comment{Check eq.~\eqref{eq:consistency}}
  \If{$\{v_i, v'_i\} \cap \mathcal{C}_i = \emptyset$}
  \State \Return \textbf{False} \Comment{Bad Guesses}
  \EndIf
  \EndFor
  \State \Return \textbf{True} \Comment{No inconsistency}
  \EndProcedure

@@

\Procedure{ReconstructPartialDifference}{$X_0, \dots, X_k$}
  \State // \emph{Statement involving $j$ must be repeated for $j=0, 1, 2, 3, 4$.}
  \State $H \gets$ LLL reduction of $G_{4,64}$
  \State $\ell \gets 14$
\For{$0 \leq w_0 < 2^{\ell}$ and $0 \leq c_0 < 2^{\ell - 1}$} \Comment{Guess least-significant bits}
\State $K_j \gets a^j w_0 + (2c_0 + 1)(a^j - 1)(a-1)^{-1} \bmod 2^{128}$ \Comment{Known part}
  \For{$0 \leq r_0, r_1, r_2, r_3, r_4 < 64$} \Comment{Guess rotations}
  \State $Y_j \gets X_j \lll r_j$ \Comment{Undo rotations}
  \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(K_j \oplus Y_j\right)[0:\ell]$ \Comment{Truncated LCG}
  \State $\widetilde{T}'_j \gets T_j \boxminus  K_i[58:64+\ell]$ \Comment{Cancel known part}
  \State $\widetilde{\Delta T'}_j = \widetilde{T'}_{j+1} \boxminus \widetilde{T}'_j$  \Comment{Difference (truncated geom. seq.)}
  \State $(\Delta U_0, \dots, \Delta U_3) \gets \left\lfloor (\Delta T'_0, \dots, \Delta T'_3) \cdot 2^{58-\ell} \cdot \widetilde H^{-1} \right\rceil \cdot \widetilde H$ \Comment{CVP}
  \State $\Delta S_0[0:64+\ell] \gets \left(K_1 \boxminus K_0\right)[0:\ell]
  + 2^{\ell} \cdot \Delta U_0[0:64]$ \Comment{Check}
  \If{$\Call{ConsistencyCheck}{\Delta_0, w_0, Y_0, X_5, \dots, X_k}$}
  \State \Return $(w_0, c, r_0, \dots, r_4, \Delta_0)$.
  \EndIf
  \EndFor
  \EndFor
  \EndProcedure
\end{algorithmic}
\caption{Partial difference reconstruction algorithm (when $c$ is unknown).}
\label{algo:unknown_1}
\end{algorithm}

The heart of the algorithm is again the reconstruction of a truncated geometric
progression. Looking at table~\ref{tab:babai}, we see that the best choice
consists in guessing 5 consecutive rotations and $\ell=14$ least-significant
bits. Therefore, \textsc{ReconstructPartialDifference} does $2^{57}$ iterations
of the inner loop, and succeeds deterministically.

\subsection{Predicting all the Rotations}

Knowing the values of $\Delta S_0[0:64+\ell]$ as well as the $\ell$
least-significant bits of $S_0$ and $c$ is sufficient to get rid the nastier
feature \textsf{PCG64}: armed with this knowledge, we can determine all the
subsequent rotations deterministically, at negligible cost, using
eq~\eqref{eq:find_rotation}. For each index $i$, it suffices to try the $64$
possible values of $r_i$; only one should satisfy
eq~\eqref{eq:find_rotation}. The complete pseudo-code is shown in
algorithm~\ref{algo:unknown_2}.

It is unlikely that several possible values of $r_i$ match: each value is
``checked'' on $\ell$ bits, so an accidental match happens with probability
$2^{\ell-6}$. The total number of lists returned by
\textsc{ReconstructRotations} then follows a binomial distribution of parameters
$2^{\ell-6}, k$. With $\ell=14$ and $k=64$, then only one rotation vector should
pass the test for $0 \leq i < 64$ on average.

@@

  %\State $x_i = x_0 + \left(\Delta S_0 \frac{a^i-1}{a-1} \bmod 2^{64+\ell}\right)[64:64+\ell]$. \Comment{$x_i = S_i[64:64+\ell]$}
  \For{$0 \leq r < 64$} \Comment{Try all rotations}
  \If{$w_i \oplus (X_i \lll r)[0:\ell] \in \left\{v_i, v'_i\right\}$} \Comment{Check eq.~\eqref{eq:find_rotation}}
  \State $\mathcal{H} \gets r \texttt{::} \mathcal{H}$ \Comment{New candidate $r_i$}
  \EndIf
  \EndFor
  \State \Return $\left\{ h\texttt{::} t~|~ h \in \mathcal{H}, t \in \mathcal{T} \right\}$ \Comment{Return $\mathcal{H} \times \mathcal{T}$}
  \EndFunction
\end{algorithmic}
\caption{Rotations and full difference reconstruction algorithm}
\label{algo:unknown_2}
\end{algorithm}

Using $X_0, X_1, \dots, X_{63}$, we recover all rotations and thus we recover
the 6 most-significant bits of $S_0, S_1, \dots, S_{63}$. This allows us to
compute the 6 most significant bits of the differences $\Delta S_i$ between
consecutive states (up to missing carries), and we are faced with the problem of
reconstructing a 128-bit geometric progression using 63 consecutive outputs
truncated to their 6 most-significant bits. There is again an unknown vector of
borrows $B$ such that $\Delta S_i[122:128] \boxplus C_i= r_{i+1} \boxminus r_i$.

Reconstructing $\Delta S_0$ from the $r_i$ is exactly the problem discussed in
section~\ref{sec:cvp_big}. This can be done by solving an instance of CVP in
dimension $63$. We use the off-the-shelf CVP solver embedded in \texttt{fplll}:
it runs in no measurable time.
